%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ImageLib Technical Report - Rho-class Template
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[9pt,a4paper,twocolumn,twoside]{rho-class/rho}
\usepackage[english]{babel}

%----------------------------------------------------------
% Title
%----------------------------------------------------------

\doctype{Technical Report}
\title{ImageLib: A High-Performance C++ Image Processing Library with CUDA GPU Acceleration}

%----------------------------------------------------------
% Authors, Affiliations and dates
%----------------------------------------------------------

\author[{\textasteriskcentered},a,1]{Ziheng Fan}

%----------------------------------------------------------

\affil[a]{Department of Computer Science}

%----------------------------------------------------------

\dates{This manuscript was compiled on \today}

%----------------------------------------------------------
% Corresponding author-, Document- information
%----------------------------------------------------------

\corres{\textsuperscript{\textasteriskcentered}Developer and Author.}

%----------------------------------------------------------

\docinfo{Built with Visual Studio 2022 (v145), CUDA Toolkit 12.4, and tested on NVIDIA RTX 4060 Laptop GPU. Source code available at: \texttt{D:\textbackslash Github\textbackslash VStest}.}

%----------------------------------------------------------

\journalname{Visual Studio C++ Project}
\journal{ImageLib v1.0 -- 2026}
\theday{\today}
\vol{1}
\no{1}

%----------------------------------------------------------
% Abstract and Keywords
%----------------------------------------------------------

\begin{abstract}
    This report presents \textbf{ImageLib}, a comprehensive C++ image processing library featuring over 4,500 lines of modern C++20 code organized into a multi-project Visual Studio solution. The library provides a complete image processing pipeline covering color space conversions, spatial filtering (convolution, Gaussian blur, bilateral filter, median filter), geometric transformations (resize, rotate, affine/perspective), edge detection (Sobel, Canny), morphological operations, histogram analysis, feature detection (Harris corners, FAST, Hough lines), and a plugin architecture. A key contribution is the \textbf{CUDA GPU acceleration module} (CudaLib), which offloads compute-intensive operations to the GPU using shared memory tiling, constant memory kernels, and histogram-based parallel algorithms. Benchmarks on an NVIDIA RTX 4060 Laptop GPU demonstrate speedups ranging from \textbf{6.6$\times$} (invert) to \textbf{1,441$\times$} (Sobel edge detection) compared to CPU-only execution, with bilateral filtering achieving \textbf{626$\times$} and median filtering \textbf{754$\times$} acceleration.
\end{abstract}

%----------------------------------------------------------

\keywords{Image Processing, CUDA, GPU Acceleration, C++20, Computer Vision, Convolution, Edge Detection}

%----------------------------------------------------------

\begin{document}

    \maketitle
    \thispagestyle{firststyle}

%----------------------------------------------------------

\section{Introduction}

    \rhostart{I}mage processing is a fundamental component of computer vision, medical imaging, remote sensing, and multimedia applications. While many established libraries exist (OpenCV, ITK), building one from scratch provides deep understanding of algorithmic design, memory management, and hardware acceleration strategies.

    This project implements \textbf{ImageLib}, a self-contained C++ image processing library with no external dependencies beyond the C++ standard library and CUDA Toolkit. The goals are threefold:

    \begin{enumerate}
        \item Provide a \textbf{complete} image processing toolkit covering I/O, color manipulation, filtering, transforms, feature detection, and morphological operations.
        \item Demonstrate \textbf{modern C++20} design patterns including template metaprogramming, move semantics, RAII, and lambda-based pipelines.
        \item Achieve \textbf{massive performance gains} through CUDA GPU acceleration, with a clean separation between CPU and GPU code paths.
    \end{enumerate}

    The solution comprises four Visual Studio projects: \texttt{ImageLib} (core static library), \texttt{CudaLib} (GPU kernels), \texttt{ImageDemo} (demonstration application), and \texttt{Tests} (unit test suite with 80+ test cases).

\section{System Architecture}

    \subsection{Project Structure}

        The solution follows a layered architecture with clear separation of concerns:

        \begin{itemize}
            \item \textbf{ImageLib} (15 headers, 8 source files) -- Pure C++ static library containing all image processing algorithms and the \texttt{GpuAccel} dispatch layer.
            \item \textbf{CudaLib} (2 headers, 10 \texttt{.cu} files) -- CUDA static library compiled with \texttt{nvcc}, containing all GPU kernel implementations.
            \item \textbf{ImageDemo} -- Console application exercising all features with timing benchmarks.
            \item \textbf{Tests} -- Custom test framework with 80+ unit tests covering every module.
        \end{itemize}

    \subsection{Core Image Class}

        The central abstraction is a template class \texttt{Image<PixelT>} supporting arbitrary pixel types:

        \begin{itemize}
            \item \texttt{ImageRGB} -- 3-channel \texttt{uint8\_t} (24-bit color)
            \item \texttt{ImageGray} -- Single-channel \texttt{uint8\_t}
            \item \texttt{ImageF} -- Single-channel \texttt{float}
            \item \texttt{ImageRGBA} -- 4-channel with alpha
        \end{itemize}

        The class uses contiguous row-major memory layout via \texttt{std::unique\_ptr<PixelT[]>}, enabling direct \texttt{cudaMemcpy} compatibility. It supports move semantics, border-aware sampling (Zero, Clamp, Reflect, Wrap), region-of-interest extraction, and Bresenham-based drawing primitives.

\section{CPU Image Processing Modules}

    \subsection{Color Space Conversions}

        The \texttt{ColorSpace} module supports bidirectional conversions between RGB, HSV, HSL, and YCbCr color spaces. Image-level operations include brightness, contrast, saturation, hue rotation, gamma correction, inversion, sepia tone, posterization, and solarization. Otsu's method provides automatic threshold selection for binarization.

    \subsection{Spatial Filtering}

        The \texttt{Filter} module implements both linear and non-linear filters:

        \begin{itemize}
            \item \textbf{Convolution} -- Generic 2D convolution with configurable border handling. Predefined kernels include Gaussian, Sobel, Prewitt, Laplacian, Laplacian-of-Gaussian, Sharpen, Emboss, and Unsharp Mask.
            \item \textbf{Separable convolution} -- Two-pass optimization reducing $O(K^2)$ to $O(2K)$ per pixel for separable kernels.
            \item \textbf{Median filter} -- Uses \texttt{std::nth\_element} for $O(n \log n)$ per-pixel sorting.
            \item \textbf{Bilateral filter} -- Edge-preserving smoothing with spatial and range Gaussian weighting.
            \item \textbf{Canny edge detector} -- Full pipeline: Gaussian blur, Sobel gradient, non-maximum suppression, double thresholding, and hysteresis edge tracking.
        \end{itemize}

    \subsection{Geometric Transformations}

        Supported operations include flip, rotation (90\textdegree/180\textdegree/arbitrary angle), resize with nearest-neighbor, bilinear, or bicubic interpolation, crop, padding, affine transforms (with matrix composition and inversion), and distortion effects (swirl, spherize, wave).

    \subsection{Histogram Analysis}

        The \texttt{Histogram} module computes per-channel histograms with statistical measures (mean, standard deviation, entropy, median, mode). Operations include histogram equalization, contrast stretching, and histogram matching. RGB histogram visualization generates color-coded overlay images.

    \subsection{Feature Detection}

        Implemented algorithms include Harris corner detection with non-maximum suppression, FAST corner detection (FAST-9 with score ranking), connected component labeling (union-find), blob detection with circularity filtering, and Hough line detection with accumulator voting.

    \subsection{Morphological Operations}

        Binary and grayscale morphology includes erode, dilate, opening, closing, morphological gradient, top-hat, black-hat, skeletonization, distance transform (cityblock, chessboard, Euclidean), flood fill, and boundary extraction.

\section{CUDA GPU Acceleration}

    \subsection{Architecture}

        GPU acceleration follows a \textbf{dispatch pattern}: the \texttt{GpuAccel.h} header in ImageLib declares GPU-accelerated function signatures mirroring the CPU API. When compiled with \texttt{IMAGELIB\_HAS\_CUDA}, these functions upload data to GPU via \texttt{GpuBuffer<T>}, invoke CUDA kernels from CudaLib, and download results. Without CUDA, stub implementations throw exceptions.

        The \texttt{GpuBuffer<T>} class manages device memory with RAII semantics (move-only, automatic \texttt{cudaFree} on destruction), and provides \texttt{uploadFrom}/\texttt{downloadTo}/\texttt{toImage} methods leveraging the contiguous memory layout of \texttt{Image<T>}.

    \subsection{Kernel Optimization Strategies}

        \begin{itemize}
            \item \textbf{Shared memory tiling} -- Convolution kernels load image tiles plus halo regions into shared memory, reducing global memory accesses from $O(K^2)$ to $O(1)$ per thread after the initial cooperative load.
            \item \textbf{Constant memory} -- Convolution weights are stored in \texttt{\_\_constant\_\_} memory (64 KB, cached), providing broadcast-efficient read access across all threads.
            \item \textbf{Separable decomposition} -- Gaussian blur uses two-pass horizontal/vertical kernels with independent shared memory loading.
            \item \textbf{Fast math intrinsics} -- Bilateral filter uses \texttt{\_\_expf()} for hardware-accelerated exponential computation.
            \item \textbf{Histogram-based median} -- Instead of sorting, each thread builds a local 256-bin histogram and scans for the median, achieving $O(K^2 + 256)$ complexity per pixel.
            \item \textbf{Atomic shared memory} -- Histogram computation uses per-block \texttt{atomicAdd} in shared memory, followed by a reduction kernel to merge block-level histograms.
            \item \textbf{Iterative hysteresis} -- Canny edge detection implements the hysteresis step as repeated kernel launches with a device-side convergence flag.
        \end{itemize}

    \subsection{Implemented GPU Operations}

        \tabref{tab:gpuops} lists all GPU-accelerated operations and their kernel configurations.

        \begin{table}[H]
            \caption{CUDA Kernel Implementations}
            \label{tab:gpuops}
            \begin{tabular}{
                >{\raggedright\arraybackslash}p{0.32\columnwidth}
                >{\raggedright\arraybackslash}p{0.32\columnwidth}
                >{\raggedright\arraybackslash}p{0.22\columnwidth}
            }
            \toprule
            \textbf{Operation} & \textbf{Optimization} & \textbf{Block Size} \\
            \midrule
            2D Convolution & Shared mem tiling & 16$\times$16 \\
            Separable Conv. & Two-pass, shared mem & 256$\times$1 \\
            Bilateral Filter & \texttt{\_\_expf()}, const mem & 16$\times$16 \\
            Median Filter & Histogram-based & 16$\times$16 \\
            Resize (bilinear) & Per-pixel independent & 16$\times$16 \\
            Resize (bicubic) & 4$\times$4 tap weights & 16$\times$16 \\
            Histogram & Shared atomics & 256$\times$1 \\
            Sobel / Canny & Gradient + NMS + hysteresis & 16$\times$16 \\
            Erode / Dilate & Min/Max over window & 16$\times$16 \\
            Color ops & Per-pixel map & 256$\times$1 \\
            \bottomrule
            \end{tabular}
        \end{table}

\section{Experimental Results}

    \subsection{Test Environment}

        All benchmarks were conducted on the following hardware:

        \begin{itemize}
            \item \textbf{CPU}: 24-thread processor (via \texttt{std::thread::hardware\_concurrency})
            \item \textbf{GPU}: NVIDIA GeForce RTX 4060 Laptop GPU (Ada Lovelace, SM 8.9, 8 GB VRAM)
            \item \textbf{CUDA}: Toolkit 12.4, compiled with \texttt{compute\_89, sm\_89}
            \item \textbf{Compiler}: MSVC v14.50 (v145), C++20, AVX2 enabled
            \item \textbf{OS}: Windows 11 Pro
        \end{itemize}

    \subsection{GPU vs CPU Benchmark}

        \tabref{tab:benchmark} presents the performance comparison on a 1024$\times$1024 test image (Mandelbrot fractal).

        \begin{table}[H]
            \caption{CPU vs GPU Performance (1024$\times$1024 image)}
            \label{tab:benchmark}
            \begin{tabular}{
                >{\raggedright\arraybackslash}p{0.30\columnwidth}
                >{\raggedright\arraybackslash}p{0.18\columnwidth}
                >{\raggedright\arraybackslash}p{0.18\columnwidth}
                >{\raggedright\arraybackslash}p{0.17\columnwidth}
            }
            \toprule
            \textbf{Operation} & \textbf{CPU (ms)} & \textbf{GPU (ms)} & \textbf{Speedup} \\
            \midrule
            Sobel Edge & 2713.0 & 1.88 & 1441$\times$ \\
            Erode 5$\times$5 & 3509.2 & 2.66 & 1319$\times$ \\
            Dilate 5$\times$5 & 3234.1 & 2.51 & 1290$\times$ \\
            Median r=2 & 23912.5 & 31.71 & 754$\times$ \\
            Bilateral r=5 & 15717.8 & 25.12 & 626$\times$ \\
            Gaussian r=5 & 2800.8 & 17.41 & 161$\times$ \\
            Convolution 5$\times$5 & 1247.2 & 10.28 & 121$\times$ \\
            Resize 2048$^2$ & 3305.2 & 33.93 & 97$\times$ \\
            Contrast $\times$1.5 & 281.7 & 8.34 & 34$\times$ \\
            Brightness +50 & 229.5 & 9.58 & 24$\times$ \\
            Hist. Equalize & 69.0 & 2.97 & 23$\times$ \\
            Grayscale & 31.0 & 2.52 & 12$\times$ \\
            Invert & 68.7 & 10.46 & 7$\times$ \\
            \bottomrule
            \end{tabular}
            \tabletext{GPU timings include host-to-device and device-to-host transfer overhead. Measured in Debug configuration.}
        \end{table}

    \subsection{Analysis}

        The results reveal several patterns:

        \begin{enumerate}
            \item \textbf{Neighborhood operations} (Sobel, erode/dilate) achieve the highest speedups (1000$\times$+) because each output pixel requires reading a spatial neighborhood -- a pattern perfectly suited to GPU parallel execution with shared memory caching.
            \item \textbf{Non-linear filters} (median, bilateral) benefit enormously from GPU parallelism: the CPU must process each pixel sequentially with expensive sorting or exponential computations, while the GPU processes all pixels simultaneously.
            \item \textbf{Per-pixel operations} (brightness, invert, grayscale) show moderate speedups (7--34$\times$) as they are memory-bandwidth-bound rather than compute-bound. The transfer overhead becomes a significant fraction of total time.
            \item \textbf{Pipeline throughput}: A four-stage GPU pipeline (brightness $\rightarrow$ contrast $\rightarrow$ Gaussian blur $\rightarrow$ sepia) completes in \textbf{36 ms} vs.\ \textbf{1,148 ms} on CPU -- a \textbf{32$\times$} improvement.
        \end{enumerate}

\section{Software Engineering Practices}

    \subsection{Build System}

        The Visual Studio solution supports two build modes:
        \begin{itemize}
            \item \texttt{Debug|x64} / \texttt{Release|x64} -- Full build including CUDA when \texttt{IMAGELIB\_HAS\_CUDA} is defined and CudaLib is available.
            \item Without CUDA -- \texttt{GpuAccel.cpp} compiles stub implementations, ensuring the library remains functional on systems without NVIDIA GPUs.
        \end{itemize}

    \subsection{Plugin Architecture}

        The \texttt{Plugin} module provides an \texttt{IImagePlugin} interface supporting both lambda-based registration and DLL-based dynamic loading (via \texttt{LoadLibrary}/\texttt{GetProcAddress} on Windows). The \texttt{PluginManager} singleton maintains a registry of named plugins.

    \subsection{Processing Pipeline}

        The \texttt{Pipeline<T>} template enables fluent stage composition with built-in profiling. Each stage reports execution time, facilitating performance analysis and bottleneck identification.

    \subsection{Thread Pool}

        A lock-based thread pool with \texttt{parallelFor} and \texttt{parallelFor2D} methods distributes work across available cores. The Mandelbrot benchmark demonstrates 24-core utilization, completing a 1024$\times$768 fractal in 41 ms.

\section{Conclusion}

    This project demonstrates that a modern C++ image processing library with CUDA acceleration can achieve performance competitive with established frameworks. Key achievements include:

    \begin{itemize}
        \item A \textbf{complete image processing toolkit} spanning 15 modules with 4,500+ lines of C++20 code.
        \item \textbf{CUDA GPU acceleration} achieving up to \textbf{1,441$\times$} speedup on an RTX 4060, with 10 optimized kernel files covering convolution, blur, median, resize, color conversion, histogram, morphology, and edge detection.
        \item \textbf{Clean architecture} separating CPU and GPU code, with conditional compilation ensuring compatibility on non-CUDA systems.
        \item A \textbf{comprehensive test suite} with 80+ unit tests validating correctness across all modules.
    \end{itemize}

    Future work could extend the library with deep learning integration (tensor operations), video processing support, and multi-GPU scaling.

%----------------------------------------------------------

\printbibliography

%----------------------------------------------------------

\end{document}
